---
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
devtools::load_all()
```

[![Build Status](https://travis-ci.org/oganm/markerGeneProfile.svg?branch=master)](https://travis-ci.org/oganm/markerGeneProfile)[![codecov](https://codecov.io/gh/oganm/markerGeneProfile/branch/master/graph/badge.svg)](https://codecov.io/gh/oganm/markerGeneProfile)


markerGeneProfile
==================
This package includes functions responsible for marker gene selection and marker gene profile estimation estimation as described in Mancarci et a. 2017. It also includes a copy of mouse brain cell type markers from the [neuroExpressoAnalysis](https://github.com/oganm/neuroExpressoAnalysis) package for convenience along with mock data for easy testing.

Table of contents
=============================
- [Installation](#installation)
- [Usage](#usage)
  - [Marker genes]

Installation
=============

Use devtools to install
```
devtools::install_github('oganm/markerGeneProfile')
```


Usage
=============

Marker genes
------------------
A marker gene list is needed in order to estimate cell type profiles. In this package, a copy of mouse brain cell type markers from [neuroExpressoAnalysis](https://github.com/oganm/neuroExpressoAnalysis), the package that summarizes the entire analysis performed in Mancarci et al. 2017 is included. If an different marker gene set is needed steps below can be followed to create one

### Sample data for marker gene selection

The package includes a sample cell type profile dataset aimed to demonstrate the minimal information required for selection of marker genes.

`mgp_sampleProfilesMeta` includes the basic metadata required for the cell type specific expression dataset.

```{r}
kable(head(mgp_sampleProfilesMeta))
```

**sampleName:** name of the samples. This needs to correspond to column names in the expression file.

**replicate:** A vector marking which samples are replicates of each other.

**PMID: ** A vector marking which samples come from the same study. Normally taking PMIDs of the papers is a good idea.

**CellType:** A vector marking the cell types that the samples represent.

**region:** The regions samples are extracted from. Only needed if region specific genes are to be selected.

**RegionToParent:** If region specific genes are to be selected and a region hierarchy is to be used, this column controls whether or not the sample should be included in the parent regions of the indicated region. If not provided it will default to `TRUE`. The name of this column is hard coded and should not be changed.

**RegionToChildren:** Same as above except it controls if the sample should be included in the children regions. If not provided it will default to `TRUE`. The name of this column is hard coded and should not be changed.

`mgp_sampleProfiles` is a sample expression data. **Gene.Symbol** column is the gene identifier that should be composed of unique IDs while the rest are sample names that corresponds to the relevant column in the metadata file. Other columns can be present before the sample data but they should not be of class `double`.

```{r}
kable(mgp_sampleProfiles)
```

`mpg_sampleRegionHiearchy` is a sample region hiearchy. It is a nested named list.

```{r}
mpg_sampleRegionHiearchy

ogbox::nametree(mpg_sampleRegionHiearchy)
```
In this example `Region 1` and `Region 2` are subsets of `All` region

### Selection of marker genes

Marker gene selection is performed using three functions: `markerCandidates`, `pickMarkers` and `rotateSelect`. By default `markerCandidates` will return files for each cell type in a region that lists the gene that are above a given silhouette and fold change thresholds. Other variables are briefly explained below but see the package documentation for in depth explanations

```{r}
markerCandidates(design = mgp_sampleProfilesMeta, # the design file
                 expression = mgp_sampleProfiles, # expression file 
                 outLoc = 'readmeDocs/quickSelection', # output directory
                 groupNames = 'CellType', # name of the column with cell types. can be a vector
                 regionNames = 'region', # name of the column with brain regions. leave NULL if no region seperation is desired
                 PMID = 'PMID', # name of the column with study identifiers
                 sampleName = 'sampleName', # name of the column with sample names
                 replicates = 'replicate', # name of the column with replicates
                 foldChangeThresh = 10, # threshold of fold change for gene selection (default is 10)
                 minimumExpression = 8, # minimum expression level that a gene can be considered a marker gene (default is 8)
                 background = 6, # background level of expression (default is 6)
                 regionHierarchy = mpg_sampleRegionHiearchy, # hierarchy of brain regions to be used
                 geneID = 'Gene.Symbol', # column name with with gene idenditifers
                 cores = 16 # number of cores to use in parallelization 
                 )
```
This creates 3 directories in the output directory 
```{r}
list.files('readmeDocs/quickSelection')
```
The `CellType` directory is a list of marker genes that disregards all region specifications (redundant with `All_CellType` in this case) while `Region 2_CellType` and `All_CellType` directories inlcude cell types from the relevant region. Note the absence of `Region 1_CellType` since that region only has a single cell type. 

```{r}
read.table('readmeDocs/quickSelection/All_CellType/Cell C') %>% kable
```

This file shows the candidate genes for cell type `Cell C` in region `All`. The first column is the gene identifier, the second is change in expression in log_2 scale and the third one is the silhouette coefficient. Note that `Gene6` is absent since its expression level was below the minimum level allowed. `markerCandidates` function does not apply a threshold for silhouette coefficient it also doesn't check to see if a gene satisfies fold change threshold for multiple genes. `pickMarkers` function does that.

```{r}
pickMarkers('readmeDocs/quickSelection/All_CellType/',
            foldChange = 10,
            silhouette = 0.5)
```
If all genes for all regions needs to be seen
```{r}
pickMarkersAll('readmeDocs/quickSelection')

```


### Better selection of marker genes

The above method is a quick way to pick markers but it does not handle bimodality in expression distribution well. To ensure robustness of the results it is better to perform multiple selections with permutations. `markerCandidates` function has variables to handle permutations for you. `rotate` controls what is the percentage of samples that should be removed every time. seed controls the random seed and is there to ensure reproducibility.

```{r, results = 'hide'}
for (i in 1:10){
    markerCandidates(design = mgp_sampleProfilesMeta, # the design file
                     expression = mgp_sampleProfiles, # expression file 
                     outLoc = file.path('readmeDocs/Rotation',i), # output directory
                     groupNames = 'CellType', # name of the column with cell types. can be a vector
                     regionNames = 'region', # name of the column with brain regions. leave NULL if no region seperation is desired
                     PMID = 'PMID', # name of the column with study identifiers
                     sampleName = 'sampleName', # name of the column with sample names
                     replicates = 'replicate', # name of the column with replicates
                     foldChangeThresh = 10, # threshold of fold change for gene selection (default is 10)
                     minimumExpression = 8, # minimum expression level that a gene can be considered a marker gene (default is 8)
                     background = 6, # background level of expression (default is 6)
                     regionHierarchy = mpg_sampleRegionHiearchy, # hierarchy of brain regions to be used
                     geneID = 'Gene.Symbol', # column name with with gene idenditifers
                     cores = 16, # number of cores to use in parallelization 
                     rotate = 0.33,
                     seed = i
    )
}
```
This creates multiple selection directories. `rotateSelect` can be used to count the number of times a gene is selected for each cell type in each region. This creates another directory similar to the output of `markerCandidates`. Again, valid markers can be acquired using `pickMarkers`

```{r, results= 'hide'}
rotateSelect(rotationOut='readmeDocs/Rotation',
                 rotSelOut='readmeDocs/RotSel',
                 cores = 16,foldChange = 1)


```
